Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation

### 2024-05-31 paper/ms.tex
more edits to intro and refs

\firstpage{3}
% Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting. In this example, there are a total of 11 paragraphs on the first page, counting the first level heading as a paragraph. The value {12} represents the number of the paragraph starting the second page. If a paragraph runs over onto the second page, include a bracket with the paragraph line number starting the second page, followed by the paragraph number in curly brackets, e.g. "\firstpage[4]{11}".

@@

The particle filter, also known as sequential Monte Carlo, serves as the foundation for various inference algorithms for POMP models.
It provides an unbiased estimate of the likelihood function  \cite{delMoral04}, enabling Bayesian inference \cite{andrieu10,chopin13} and maximum likelihood inference \citep{ionides06-pnas,ionides15}.
An attractive feature of basic particle filter algorithms, also known as boostrap filters, is that they do not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
This plug-and-play property is useful for scientific applications \cite{he10}.
Some plug-and-play methods depend on the construction of low-dimensional summary statistics \cite{wood10,toni09}, sacrificing statistical efficiency for computational convenience.
However, particle filter inference methods based on the likelihood are statistically efficient, up to Monte Carlo error.
[MAYBE ADD SOME REFERENCES HERE TO DOCUMENT SUCCESSES OF IF2, SINCE WE ARE COMPARING WITH THAT AND BUILDING ON IT]
This Monte Carlo error can be non-negligible.
For example, though in practice the algorithm of Ionides et al. \cite{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood.
The issue becomes increasingly problematic as the dataset size and model complexity increase, necessitating a new approach.

>>>>

\firstpage[4]{4}
% Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting. In this example, there are a total of 11 paragraphs on the first page, counting the first level heading as a paragraph. The value {12} represents the number of the paragraph starting the second page. If a paragraph runs over onto the second page, include a bracket with the paragraph line number starting the second page, followed by the paragraph number in curly brackets, e.g. "\firstpage[4]{11}".

@@

The particle filter, also known as sequential Monte Carlo, serves as the foundation for various inference algorithms for POMP models.
It provides an unbiased estimate of the likelihood function  \cite{delMoral04}, enabling Bayesian inference \cite{andrieu10,chopin13} and maximum likelihood inference \citep{ionides06-pnas,ionides15}.
An attractive feature of basic particle filter algorithms, also known as boostrap filters, is that they do not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.
This plug-and-play property is useful for scientific applications \cite{he10}.
Some plug-and-play methods depend on the construction of low-dimensional summary statistics \cite{wood10,toni09}, sacrificing statistical efficiency for computational convenience.
However, particle filter inference methods based on the likelihood are statistically efficient, up to Monte Carlo error.
Plug-and-play methods are often called likelihood-free \cite{owen15}, and it may be counter-intuitive that likelihood-based inference is possible using likelihood-free methods. 
However, iterated filtering algorithms have shown that this is practical in various scientific investigations \cite{king08,blake14,pons-salort18,subramanian21,fox22,drake23}.
Nevertheless, the Monte Carlo variability arising in iterated filtering applications becomes increasingly non-negligible as the size of the data and the complexity of the model increases.

For example, though in practice the algorithm of Ionides et al. \cite{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood.
The issue becomes increasingly problematic as the dataset size and model complexity increase, necessitating a new approach.
